---
title: "Day 1A"
format:
  html:
    css: ../styles.css
    highlight-style: github
---

![](../img/teaching_2780E3.svg){.hero}

We begin with an overview of the workshop, an introduction to R and RStudio, and the basics of programming. The latter will cover basic math operations, creating and naming objects, and calling functions.

## Slides

<iframe class="slide-deck" src="./Slides/1A_Slides.html" title="Day 1A Slideshow">
</iframe>

[Click here to view the slides in their own window.](./Slides/1A_Slides.html)

## Practice 1

Read what the user wanted to do and look at the command they sent to R. Fix the errors in the commands and calculate the answers.

a)  I want to add 20 and 0.2 and divide that sum by 4.1. But I get an error when I try:<br />`[20 + .2] \ 4.1`
b)  I want to multiply 6 by the sum of 2 and 1,300. But I get an error when I try:<br />`6 x (2 + 1,300)`

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Answer (a)**

Change the brackets `[]` to parentheses `()` because parentheses `()` are the order of operations operator in R. Also, change the backslash `\` to a forward slash `/` because forward slash `/` is the division operator in R.

```{r}
#| comment: "#>"
#| collapse: true

(20 + .2) / 4.1
```

**Answer (b)**

Change the `x` to an asterisk `*` because asterisk `*` is the multiplication operator in R. Also, remove the comma `,` from `1,300` because R doesn't use commas to delimit digits in large numbers.

```{r}
#| comment: "#>"
#| collapse: true

6 * (2 + 1300)
```

</blockquote></details></p>


## Practice 2

Which of the following commands do you think will create an error in R? Why?

a) `score@T1 <- 3.2`
b) `score_at_T1 <- 3.2`
c) `score at T1 <- 3.2`
d) `1_score <- 3.2`
e) `ScoreAtTime1 <- 3.2`

Make your best guesses just by looking at the commands and then check your guesses by running the commands in R.

<p><details>
<summary>Click here for the answer key</summary>

> The following commands will create an error in R: **a** (`@` is not allowed in an object name), **c** (spaces are not allowed in an object name), and **d** (the first character in an object name cannot be a number).

</details></p>



## Practice 3

a) Use the `cos()` function to calculate the cosine of 3 and assign it to a variable called `y`. Print the value of `y` to the console. 
b) Use a function to round `y` to the nearest whole number.
c) Use a function to round `y` to two decimal places.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Answer (a)**
```{r}
#| collapse: true

y <- cos(3)
y
```

**Answer (b)**
```{r}
#| collapse: true

round(y)
```

**Answer (c)**
```{r}
#| collapse: true

round(y, digits = 2)
```


</blockquote></details></p>



## Practice 4

What do you expect the result to be for the following commands?

a) `sqrt(4 + 5)`
b) `sqrt(4) + 5`

Make your best guesses just by looking at the commands and then check your guesses by running the commands in R.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Answer (a)**
```{r}
#| collapse: true

sqrt(4 + 5)

```

Add then square root, same as sqrt(9)

**Answer (b)**
```{r}
#| collapse: true

sqrt(4) + 5 
```

Square root then add, same as 2 + 5

</blockquote></details></p>


## Readings

  -   ["Chapter 3: Workflow - Basics"](https://r4ds.hadley.nz/workflow-basics.html) from *R4DS (2E)*
  -   ["Chapter 7: Workflow - Scripts and Projects"](https://r4ds.hadley.nz/workflow-scripts.html) from *R4DS (2E)*

## Fun Stuff

### "The Literal Genie" 

What it feels like to talk to R sometimes...

{{< video https://www.youtube-nocookie.com/embed/58l7ZqoU0a0 >}}


::: {.text-center}
<a href="day1B.html" class="btn btn-primary mt-5" role="button" >Continue to Day 1B Materials&nbsp;&nbsp;&raquo;</a>
:::
