---
title: "Day 2A"
format:
  html:
    css: ../styles.css
    highlight-style: github
---

![](../img/statistics_2780E3.svg){.hero}

We resume on Day 2 with a discussion of special data types (for categorical data and missing values), data summarization and exploration, and basic statistical inference. The latter will include various types of correlation, two-sample tests for comparing two groups, and oneway tests for comparing multiple groups.

## Slides

<iframe class="slide-deck" src="./Slides/2A_Slides.html" title="Day 2A Slideshow">
</iframe>

[Click here to view the slides in their own window.](./Slides/2A_Slides.html)

## Data Files

- <a href="yearspubs.csv" download="yearspubs.csv">yearspubs</a> (professors' salaries by seniority, publications, etc.)
- <a href="salaries.csv" download="salaries.csv">salaries</a> (professors' salaries by rank, discipline, etc.)
- <a href="interpersonal.csv" download="interpersonal.csv">interpersonal</a> (patients' interpersonal problems before and after treatment)
- <a href="elicitation.csv" download="elicitation.csv">elicitation</a> (participants' emotions elicited during different tasks)

## Practice

## Question 1.

a) Download the `yearspubs` dataset from the link above and read it into R.

b) Estimate the correlation between academics' number of publications (`n_pubs`) and number of citations (`n_cites`) in this sample.

c) Compare the average number of citations between male and female academics using a Welch t-test. Which group had more and was the difference significant?

d) Estimate the effect size of this difference.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Answer (a)**

```{r}
#| comment: "#>"
#| message: false
#| collapse: true

library(tidyverse)
yearspubs <- read_csv("yearspubs.csv")
```


**Answer (b)**

```{r}
#| message: false
#| warning: false

library(easystats)
correlation(yearspubs)
```

The correlation between number of publications and citations is 0.34.

**Answer (c)**

```{r}
#| eval: false

results <- t.test(
  formula = n_cites ~ sex,
  data = yearspubs
)
model_parameters(results)
```

```{r}
#| echo: false

results <- t.test(
  formula = n_cites ~ sex,
  data = yearspubs
)
model_parameters(results, effectsize_type = "d") |> print_md()
```

Group 2 (Males) had more citations (42.46) than Group 1 (Females) had (37.33) but this difference was not significant ($p=.232$).

**Answer (d)**

The effect size was $d=-0.31$ but the difference was not significant $(p=.232)$.

</blockquote></details></p>


## Question 2

Using the `mpg` tibble from the \{tidyverse\} package, statistically test whether cars with the three different drive trains (`drv`) get significantly different city fuel efficiency (`cty`) on average. Calculate the effect size for this difference.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
#| eval: false

results <- aov(formula = cty ~ drv, data = mpg)
model_parameters(results, effectsize_type = "eta")
```

```{r}
#| echo: false

results <- aov(formula = cty ~ drv, data = mpg)
model_parameters(results, effectsize_type = "eta") |> print_md()
```

The effect size was $\eta^2=0.45$ and the difference was significant $(p<.001)$.

</blockquote></details></p>


## Question 3.

Did you bring your own data? If so, load it into R and explore it using the tools we learned this week. Wrangle it a bit and maybe do some basic statistics on it, estimating correlations and comparing group means.

## Readings

  -   [Chapter 15: Factors](https://r4ds.had.co.nz/factors.html) from *R4DS (2E)*
  -   [Part IV: Statistical Theory](https://learningstatisticswithr.com/book/part-iv-statistical-theory.html) from *LSWR*
  -   [Part V: Statistical Tools](https://learningstatisticswithr.com/book/index.html) from *LSWR*
  -   [Correlation Types](https://easystats.github.io/correlation/articles/types.html) from *correlation* package website
  -   [Estimate Means](https://easystats.github.io/modelbased/articles/estimate_means.html) from *modelbased* package website
  -   [Estimate Contrasts](https://easystats.github.io/modelbased/articles/estimate_contrasts.html) from *modelbased* package

::: {.text-center}
<a href="day2B.html" class="btn btn-primary mt-5" role="button" >Continue to Day 2B Materials&nbsp;&nbsp;&raquo;</a>
:::
